---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```
# FastJM

<!-- badges: start -->
[![R-CMD-check](https://github.com/shanpengli/FastJM/workflows/R-CMD-check/badge.svg)](https://github.com/shanpengli/FastJM/actions)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/FastJM)](https://cran.r-project.org/package=FastJM)
[![](https://cranlogs.r-pkg.org/badges/grand-total/FastJM)](https://cran.r-project.org/package=FastJM)
[![CRAN_time_from_release](https://www.r-pkg.org/badges/ago/FastJM)](https://cran.r-project.org/package=FastJM)
[![CRAN_Status_Badge_version_last_release](https://www.r-pkg.org/badges/version-last-release/FastJM)](https://cran.r-project.org/package=FastJM)
<!-- badges: end -->

The `FastJM` package implement efficient computation of semi-parametric joint model of longitudinal and competing risks data.

# Example

The `FastJM` package comes with several simulated datasets. To fit a joint model, we use `jmcs` function. 

```{r, eval=TRUE}
require(FastJM)
data(ydata)
data(cdata)
fit <- jmcs(ydata = ydata, cdata = cdata, 
            long.formula = response ~ time + gender + x1 + race, 
            surv.formula = Surv(surv, failure_type) ~ x1 + gender + x2 + race, 
            random =  ~ time| ID)
fit
```

The `FastJM` package can make dynamic prediction given the longitudinal history information. Below is a toy example for competing risks data. Conditional cumulative incidence probabilities for each failure will be presented.

```{r, eval=TRUE}
ND <- ydata[ydata$ID %in% c(419, 218), ]
ID <- unique(ND$ID)
NDc <- cdata[cdata$ID  %in% ID, ]
survfit <- survfitjmcs(fit, 
                       ynewdata = ND, 
                       cnewdata = NDc, 
                       u = seq(3, 4.8, by = 0.2), 
                       method = "GH",
                       obs.time = "time")
survfit
```

To assess the prediction accuracy of the fitted joint model, we may run `PEjmcs` to calculate the Brier score. 
```{r, eval=TRUE}
## evaluate prediction accuracy of fitted joint model using cross-validated Brier Score
PE <- PEjmcs(fit, seed = 100, landmark.time = 3, horizon.time = c(3.6, 4, 4.4), 
             obs.time = "time", method = "GH", 
             quadpoint = NULL, maxiter = 1000, n.cv = 3, 
             survinitial = TRUE)
summary(PE, error = "Brier")
```

An alternative to assess the prediction accuracy is to run `MAEQjmcs` to calculate the prediction error by comparing the predicted and empirical risks stratified on different risk groups based on quantile of the predicted risks.
```{r, eval=TRUE}
## evaluate prediction accuracy of fitted joint model using cross-validated Brier Score
MAEQ <- MAEQjmcs(fit, seed = 100, landmark.time = 3, horizon.time = c(3.6, 4, 4.4), 
                 obs.time = "time", method = "GH", 
                 quadpoint = NULL, maxiter = 1000, n.cv = 3, 
                 survinitial = TRUE)
summary(MAEQ, digits = 3)
```
