---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```
# FastJM

<!-- badges: start -->
[![R-CMD-check](https://github.com/shanpengli/FastJM/workflows/R-CMD-check/badge.svg)](https://github.com/shanpengli/FastJM/actions)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/FastJM)](https://cran.r-project.org/package=FastJM)
[![](https://cranlogs.r-pkg.org/badges/grand-total/FastJM)](https://cran.r-project.org/package=FastJM)
[![CRAN_time_from_release](https://www.r-pkg.org/badges/ago/FastJM)](https://cran.r-project.org/package=FastJM)
[![CRAN_Status_Badge_version_last_release](https://www.r-pkg.org/badges/version-last-release/FastJM)](https://cran.r-project.org/package=FastJM)
<!-- badges: end -->

The `FastJM` package implement efficient computation of semi-parametric joint model of longitudinal and competing risks data.

# Example

The `FastJM` package comes with several simulated datasets. To fit a joint model, we use `jmcs` function. 

```{r, eval=TRUE}
require(FastJM)
data(ydata)
data(cdata)
fit <- jmcs(ydata = ydata, cdata = cdata, 
            long.formula = response ~ time + gender + x1 + race, 
            surv.formula = Surv(surv, failure_type) ~ x1 + gender + x2 + race, 
            random =  ~ time| ID)
fit
```

The `FastJM` package can make dynamic prediction given the longitudinal history information. Below is a toy example for competing risks data. Conditional cumulative incidence probabilities for each failure will be presented.

```{r, eval=TRUE}
ND <- ydata[ydata$ID %in% c(419, 218), ]
ID <- unique(ND$ID)
NDc <- cdata[cdata$ID  %in% ID, ]
survfit <- survfitjmcs(fit, 
                       ynewdata = ND, 
                       cnewdata = NDc, 
                       u = seq(3, 4.5, by = 0.3), 
                       M = 100,
                       seed = 100)
survfit
```
Plot the cumulative incidence function for each failure with the historical longitudinal observations.

```{r}
oldpar <- par(mfrow = c(2, 2), mar = c(5, 4, 4, 4))
plot(survfit, estimator = "both", include.y = TRUE)
par(oldpar)
```


