---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# FastJM

<!-- badges: start -->
[![R-CMD-check](https://github.com/shanpengli/FastJM/workflows/R-CMD-check/badge.svg)](https://github.com/shanpengli/FastJM/actions)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/FastJM)](https://cran.r-project.org/package=FastJM)
[![](https://cranlogs.r-pkg.org/badges/grand-total/FastJM)](https://cran.r-project.org/package=FastJM)
[![CRAN_time_from_release](https://www.r-pkg.org/badges/ago/FastJM)](https://cran.r-project.org/package=FastJM)
[![CRAN_Status_Badge_version_last_release](https://www.r-pkg.org/badges/version-last-release/FastJM)](https://cran.r-project.org/package=FastJM)
<!-- badges: end -->

The `FastJM` package implements efficient computation of semi-parametric joint model of longitudinal and competing risks data. To view a brief guide on the purpose and use of this package, please refer to our [introductory video](https://youtu.be/sspYjUATICM?si=idTbVgT5DswN-yhe).

# Examples

## Single-biomarker joint model (`jmcs`)

The `FastJM` package comes with several simulated datasets. To fit a joint model, we use `jmcs` function. In the example below, we are using the following built-in data sets:

- ydata: longitudinal data for a **single** biomarker per patient
- cdata: competing risks time-to-event data per patient

```{r}
require(FastJM)
require(survival)
data(ydata)
data(cdata)
fit <- jmcs(ydata = ydata, cdata = cdata, 
            long.formula = response ~ time + gender + x1 + race, 
            surv.formula = Surv(surv, failure_type) ~ x1 + gender + x2 + race, 
            random =  ~ time| ID)
fit
```

The `FastJM` package can make dynamic prediction given the longitudinal history information. Below is a toy example for competing risks data. Conditional cumulative incidence probabilities for each failure will be presented.

```{r}
ND <- ydata[ydata$ID %in% c(419, 218), ]
ID <- unique(ND$ID)
NDc <- cdata[cdata$ID  %in% ID, ]
survfit <- survfitjmcs(fit, 
                       ynewdata = ND, 
                       cnewdata = NDc, 
                       u = seq(3, 4.8, by = 0.2), 
                       method = "GH",
                       obs.time = "time")
survfit
```

To assess the prediction accuracy of the fitted joint model, we may run `PEjmcs` to calculate the Brier score. 
```{r}
## evaluate prediction accuracy of fitted joint model using cross-validated Brier Score
PE <- PEjmcs(fit, seed = 100, landmark.time = 3, horizon.time = c(3.6, 4, 4.4), 
             obs.time = "time", method = "GH", 
             quadpoint = NULL, maxiter = 1000, n.cv = 3, 
             survinitial = TRUE)
summary(PE, error = "Brier")
```

An alternative to assess the prediction accuracy is to run `MAEQjmcs` to calculate the prediction error by comparing the predicted and empirical risks stratified on different risk groups based on quantile of the predicted risks.

```{r}
## evaluate prediction accuracy of fitted joint model using cross-validated mean absolute prediction error
MAEQ <- MAEQjmcs(fit, seed = 100, landmark.time = 3, horizon.time = c(3.6, 4, 4.4), 
                 obs.time = "time", method = "GH", 
                 quadpoint = NULL, maxiter = 1000, n.cv = 3, 
                 survinitial = TRUE)
summary(MAEQ, digits = 3)
```

We may also calculate the area under the ROC curve (AUC) to assess the discrimination measure of joint models.
```{r}
## evaluate prediction accuracy of fitted joint model using cross-validated mean AUC
AUC <- AUCjmcs(fit, seed = 100, landmark.time = 3, horizon.time = c(3.6, 4, 4.4),
               obs.time = "time", method = "GH",
               quadpoint = NULL, maxiter = 1000, n.cv = 3, metric = "AUC")
summary(AUC, digits = 3)
```

Alternatively, we can also calculate concordance index (Cindex) as another discrimination measure.
```{r}
## evaluate prediction accuracy of fitted joint model using cross-validated mean Cindex
Cindex <- AUCjmcs(fit, seed = 100, landmark.time = 3, horizon.time = c(3.6, 4, 4.4),
               obs.time = "time", method = "GH",
               maxiter = 1000, n.cv = 3, metric = "Cindex")
summary(Cindex, digits = 3)
```

## Multi-biomarker Joint Model (`mvjmcs`)

To fit a joint model with multiple longitudinal outcomes and competing risks, we can use the `mvjmcs` function. In the example below, we are using the following built-in data sets:

- mvydata: longitudinal data for **multiple** biomarkers per patient
- mvcdata: competing risks time-to-event data per patient

```{r, eval=TRUE}
data(mvydata)
data(mvcdata)
library(FastJM)
mvfit <- mvjmcs(ydata = mvydata, cdata = mvcdata,
              long.formula = list(Y1 ~ X11 + X12 + time,
                                  Y2 ~ X11 + X12 + time),
              random = list(~ time | ID,
                            ~ 1 | ID),
              surv.formula = Surv(survtime, cmprsk) ~ X21 + X22, 
              maxiter = 1000, opt = "optim",
              tol = 1e-3, print.para = FALSE)
mvfit
```

We can extract the components of the model as follows:

```{r, eval=TRUE}
# Longitudinal fixed effects
fixef(mvfit, process = "Longitudinal")
summary(mvfit, process = "Longitudinal")

# Survival fixed effects
fixef(mvfit, process = "Event")
summary(mvfit, process = "Event")

# Random effects for first few subjects
head(ranef(mvfit))
```

The `FastJM` package can now make dynamic prediction in the presence of multiple longitudinal outcomes. Below is a toy example for competing risks data. Conditional cumulative incidence probabilities for each failure will be presented.

```{r}
require(dplyr)
set.seed(08252025)
sampleID <- sample(mvcdata$ID, 5, replace = FALSE)

subcdata <- mvcdata %>%
  dplyr::filter(ID %in% sampleID)

subydata <- mvydata %>%
  dplyr::filter(ID %in% sampleID)

### Set up a landmark time of 4.75 and make predictions at time u
survmvfit <- survfitmvjmcs(mvfit, seed = 100, ynewdata = subydata, cnewdata = subcdata,
                         u = c(7, 8, 9), Last.time = 4.75, obs.time = "time")

survmvfit

```
Currently, validation features (e.g., survfitjmcs, PEjmcs, AUCjmcs) are implemented for models of class jmcs. Extension to mvjmcs is under active development and will be available later this year.

### Simulate Data (Optional)

In order to create simulated data for `mvjmcs`, we can use the `simmvJMdata` function, which creates longitudinal and survival data as a nested list (which are unpacked the this example). When first calling the function, it provides censoring and risk rates. 

```{r}
# Simulate data
  sim <- simmvJMdata(seed = 100, N = 50) # returns list of cdata and ydata for a sample size of 50
  c_data <- sim$mvcdata # survival-side data, one row per ID
  y_data <- sim$mvydata # longitudinal measurements (multiple rows per ID)
```
Below is the simulated longitudinal data for **multiple** biomarkers, wherein Y1 and Y2 represent our biomarkers and X11 and X12 represent measurement-level predictors for the longitudinal submodel. 

```{r}
head(y_data)
```

Below is the simulated survival data wherein X21 and X22 represent patient-level predictors for the survival model.  

```{r}
head(c_data)
```
