---
title: "Joint Models for Longitudinal Measurements and Competing Risks Failure Time Data"
author: "Shanpeng Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FastJM Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Joint analysis of longitudinal outcomes and survival times has gained a lot of attention in recent years. There have been extended to handle competing risks for continuous outcomes. This vignette offers a brief introduction to the R package FastJM, which implements the methods proposed to deal with such joint models, and two competing risks are assumed. The data sets for generating the initial values are provided.

## Examples

We first load the package and the data.

```{r}
require(FastJM)
set.seed(123)
data(ydata)
data(cdata)
```
We have ydata as a longitudinal dataset with 3067 rows, 1000 subjects, and 4 variables, and cdata as a survival dataset with 1000 rows and 5 variables. 

Next let us fit a joint model using the following arguments.
```{r}

fit <- jmcs(ydata = ydata, cdata = cdata,
                    long.formula = response ~ time + x1,
                    surv.formula = Surv(surv, failure_type) ~ x1 + x2,
                    random = ~ time|ID)
```

Look at the model fit
```{r}
fit
```
The resulting table contains three parts, the fixed effects in longitudinal model, survival model and random effects. It gives the estimated parameters in the first column, standard error in the second column, 95\% confidence interval and p-value for these parameters in the third and forth columns. The estimated covariance matrix of random effects is given as sigma_b11, sigma_b12, and sigma_b22.



